---
title: "Analysis pipeline: Joint models quantify associations between T-cell kinetics and allo-immunological events after allogeneic stem cell transplantation and subsequent donor lymphocyte infusion"
author: "Edouard Bonneville and Eva Koster"
output: github_document
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

# Pipeline

The analysis pipeline for this work can be regenerated by rendering this file,

```{r, eval = FALSE}
rmarkdown::render("_targets.Rmd")
```

The pipeline can then be run using,

```{r, eval = FALSE}
tar_make()
```

The complete pipeline can be visualised using,

```{r, eval = FALSE}
tar_visnetwork()
```

# Setup

We first load the `targets` package and remove the potentially outdated workflow.

```{r}
library("targets")
library("tarchetypes")
library("future")
library("future.callr")
library("here")
tar_unscript()
```

We now define shared global options across our workflow and load R functions from the `R` folder.

```{targets globals, tar_globals = TRUE}
# Use capsule/renv eventually docker (!!)
options(tidyverse.quiet = TRUE)

# Global objects
cell_lines <- list("cell_line" = paste0("CD", c(3, 4, 8)))

# Load packages and functions necessary for pipeline
source(here::here("packages.R"))
source(here::here("data-raw/prepare-raw-data.R"))
functions <- list.files(here("R"), full.names = TRUE)
invisible(lapply(functions, source))
rm("functions")

# Run whole pipeline in parallel
plan(callr)

# Other pipeline options
tar_option_set(error = "continue")
```

# Data preparation

* First, we pre-process the raw data

```{targets process-raw-data}
list(
  tar_target(
    lymphocytes_raw,
    data.table(readRDS(here("data-raw/2022-02-24_v9/lymphocytes.rds")))
  ),
  tar_target(
    variables_raw,
    data.table(readRDS(here("data-raw/2022-02-24_v9/variables.rds")))
  ),
  tar_target(dat_merged, prepare_raw_data(lymphocytes_raw, variables_raw))
) 
```

* We add reference values for the different cell lines, for further plotting later in the pipeline

```{targets reference_values, tar_simple = TRUE}
cbind.data.frame(
  "cell_type" = c("CD3", "CD4", "CD8", "NK", "CD19"),
  "lower_limit" = c(860, 560, 260, 40, 60),
  "upper_limit" = c(2490, 1490, 990, 390, 1000)
)
```

* Next, we prepare the long and wide datasets for the alloSCT and post DLI models

```{targets prepare-data}
list(
  tar_target(
    NMA_preDLI_datasets,
    get_preDLI_datasets(
      dat_merged[TCD2 %in% c("NMA RD: ALT", "UD: ALT + ATG")], 
      admin_cens = 6
    )
  ),
  tar_target(
    NMA_postDLI_datasets,
    get_postDLI_datasets(
      dat_merged = dat_merged[TCD2 %in% c("NMA RD: ALT", "UD: ALT + ATG")],
      admin_cens_dli = 3
    )
  )
)
```

# Post alloSCT models (model I)

* Current value

```{targets preDLI-current-value}
list(
  # -- Pre-DLI Cox model
  tar_target(
    preDLI_cox,
    run_preDLI_cox(
      form = Surv(time, status) ~
        ATG.1 + hirisk.1 + # GVHD
        ATG.2 + hirisk.2 + # Relapse
        ATG.3 + # NRF other
        strata(trans),
      dat_wide = NMA_preDLI_datasets$wide
    )
  ),

  # -- Longitudinal models - correlated reffs, fixed intercept
  tar_map(
    values = cell_lines,
    tar_target(
      preDLI_long_corr,
      run_preDLI_longitudinal(
        cell_line = paste0(cell_line, "_abs_log"),
        form_fixed = "ns(intSCT2_7, 3) * hirisk * ATG + CMV_PD",
        form_random = ~ 0 + ns(intSCT2_7, 3) | IDAA,
        dat = NMA_preDLI_datasets$long
      )
    )
  ),

  # Not many events for we lower default flexibility of baseline hazard (from 5 to 3 internal knots)
  tar_target(preDLI_basehaz_knots, 3L),

  # -- Joint models, correlated reffs
  tar_target(
    preDLI_JM_value_corr_CD3,
    jointModel(
      lmeObject = preDLI_long_corr_CD3,
      survObject = preDLI_cox,
      CompRisk = TRUE,
      method = "spline-PH-aGH",
      timeVar = "intSCT2_7",
      parameterization = "value",
      interFact = list("value" = ~ strata(trans) - 1),
      iter.qN = 1000L,
      lng.in.kn = preDLI_basehaz_knots,
      numeriDeriv = "cd",
      eps.Hes = 1e-04
    )
  ),
  tar_target(
    preDLI_JM_value_corr_CD4,
    jointModel(
      lmeObject = preDLI_long_corr_CD4,
      survObject = preDLI_cox,
      CompRisk = TRUE,
      method = "spline-PH-aGH",
      timeVar = "intSCT2_7",
      parameterization = "value",
      interFact = list("value" = ~ strata(trans) - 1),
      iter.qN = 1000L,
      lng.in.kn = preDLI_basehaz_knots,
      numeriDeriv = "cd",
      eps.Hes = 1e-04
    )
  ),
  tar_target(
    preDLI_JM_value_corr_CD8,
    jointModel(
      lmeObject = preDLI_long_corr_CD8,
      survObject = preDLI_cox,
      CompRisk = TRUE,
      method = "spline-PH-aGH",
      timeVar = "intSCT2_7",
      parameterization = "value",
      interFact = list("value" = ~ strata(trans) - 1),
      iter.qN = 1000L,
      lng.in.kn = preDLI_basehaz_knots,
      numeriDeriv = "cd",
      eps.Hes = 1e-04,
      verbose = TRUE
    )
  )
)
```

* Time-dependent slopes

```{targets preDLI-slopes}
list(
  # Store the deriv form (since same for all mods)
  tar_target(
    derivForm_preDLI,
    list(
      fixed = ~ 0 + dns(intSCT2_7, 3) +
        dns(intSCT2_7, 3):as.numeric(hirisk == "yes") +
        dns(intSCT2_7, 3):as.numeric(ATG == "ALT+ATG") +
        dns(intSCT2_7, 3):as.numeric(hirisk == "yes"):as.numeric(ATG == "ALT+ATG"),
      random = ~ 0 + dns(intSCT2_7, 3),
      indFixed = c(2:4, 8:13, 15:17),
      indRandom = c(1:3)
    )
  ),
  tar_target(
    preDLI_JM_both_corr_CD3,
    update(
      preDLI_JM_value_corr_CD3,
      lmeObject = preDLI_long_corr_CD3,
      survObject = preDLI_cox,
      parameterization = "both",
      interFact = list("value" = ~ strata(trans) - 1, "slope" = ~ strata(trans) - 1),
      derivForm = derivForm_preDLI,
      lng.in.kn = preDLI_basehaz_knots
    )
  ),
  # This one gets stuck local minima.. update iter.EM to 250
  tar_target(
    preDLI_JM_both_corr_CD4,
    update(
      preDLI_JM_value_corr_CD4,
      lmeObject = preDLI_long_corr_CD4,
      survObject = preDLI_cox,
      parameterization = "both",
      interFact = list("value" = ~ strata(trans) - 1, "slope" = ~ strata(trans) - 1),
      derivForm = derivForm_preDLI,
      lng.in.kn = preDLI_basehaz_knots,
      iter.EM = 250L,
      verbose = TRUE
    )
  ),
  tar_target(
    preDLI_JM_both_corr_CD8,
    update(
      preDLI_JM_value_corr_CD8,
      lmeObject = preDLI_long_corr_CD8,
      survObject = preDLI_cox,
      parameterization = "both",
      interFact = list("value" = ~ strata(trans) - 1, "slope" = ~ strata(trans) - 1),
      derivForm = derivForm_preDLI,
      lng.in.kn = preDLI_basehaz_knots
    )
  )
)
```

* Current value and interaction

```{targets preDLI-interaction}
list(
  tar_target(
    preDLI_JM_value_corr_CD4_inter,
    update(
      preDLI_JM_value_corr_CD4,
      lmeObject = preDLI_long_corr_CD4,
      survObject = tweak_preDLI_modmat(preDLI_cox),
      parameterization = "value",
      lng.in.kn = preDLI_basehaz_knots,
      interFact = list(
        "value" = ~ trans1 + trans1:ATG.1 + trans2 + trans2:ATG.2 +
          # omit trans3 interaction since not interesting, could also do for slope models
          # i.e. include slope only relapse and gvhd
          trans3 - 1
      ),
      verbose = TRUE
    )
  ),
  tar_target(
    preDLI_JM_value_corr_CD8_inter,
    update(
      preDLI_JM_value_corr_CD8,
      lmeObject = preDLI_long_corr_CD8,
      survObject = tweak_preDLI_modmat(preDLI_cox),
      parameterization = "value",
      lng.in.kn = preDLI_basehaz_knots,
      interFact = list(
        "value" = ~ trans1 + trans1:ATG.1 + trans2 + trans2:ATG.2 + trans3 - 1
      ),
      verbose = TRUE
    )
  ),
  tar_target(
    preDLI_JM_value_corr_CD3_inter,
    update(
      preDLI_JM_value_corr_CD3,
      lmeObject = preDLI_long_corr_CD3,
      survObject = tweak_preDLI_modmat(preDLI_cox),
      parameterization = "value",
      lng.in.kn = preDLI_basehaz_knots,
      interFact = list(
        "value" = ~ trans1 + trans1:ATG.1 + trans2 + trans2:ATG.2 + trans3 - 1
      ),
      iter.EM = 250L,
      verbose = TRUE
    )
  )
)
```

# Post DLI models (model II)

```{targets postDLI-value}
list(
  # -- Longitudinal models - correlated reffs, random intercept + slope
  tar_map(
    values = cell_lines,
    tar_target(
      postDLI_long_corr,
      run_preDLI_longitudinal(
        cell_line = paste0(cell_line, "_abs_log"),
        form_fixed = "ns(intDLI1, 2) * ATG + CMV_PD",
        form_random = ~ ns(intDLI1, 2) | IDAA,
        dat = NMA_postDLI_datasets$long
      )
    )
  ),

  # -- Post-DLI cox:
  tar_target(
    postDLI_cox,
    run_postDLI_cox(
      Surv(time, status) ~ ATG.1 + strata(trans), # ATG.2 wanted but not enough events
      dat_wide = NMA_postDLI_datasets$wide
    )
  ),

  # 3L used pre-DLI, consider now 2L since only on portion of follow-up
  tar_target(postDLI_basehaz_knots, 2L),

  # -- Joint models, correlated reffs
  tar_target(
    postDLI_JM_corr_CD3,
    jointModel(
      lmeObject = postDLI_long_corr_CD3,
      survObject = postDLI_cox,
      CompRisk = TRUE,
      parameterization = "value",
      interFact = list("value" = ~ strata(trans) - 1),
      method = "spline-PH-aGH",
      timeVar = "intDLI1",
      lng.in.kn = postDLI_basehaz_knots,
      iter.EM = 1000L,
      numeriDeriv = "cd",
      eps.Hes = 1e-04,
      verbose = TRUE
    )
  ),
  tar_target(
    postDLI_JM_corr_CD4,
    jointModel(
      lmeObject = postDLI_long_corr_CD4,
      survObject = postDLI_cox,
      CompRisk = TRUE,
      parameterization = "value",
      interFact = list("value" = ~ strata(trans) - 1),
      method = "spline-PH-aGH",
      timeVar = "intDLI1",
      lng.in.kn = postDLI_basehaz_knots,
      iter.EM = 1000L,
      numeriDeriv = "cd",
      eps.Hes = 1e-04,
      verbose = TRUE
    )
  ),
  tar_target(
    postDLI_JM_corr_CD8,
    jointModel(
      lmeObject = postDLI_long_corr_CD8,
      survObject = postDLI_cox,
      CompRisk = TRUE,
      parameterization = "value",
      interFact = list("value" = ~ strata(trans) - 1),
      method = "spline-PH-aGH",
      timeVar = "intDLI1",
      lng.in.kn = postDLI_basehaz_knots,
      iter.EM = 1000L,
      numeriDeriv = "cd",
      eps.Hes = 1e-04,
      verbose = TRUE
    )
  )
)
```

# Manuscript figures

Add here targets for:

- Statistical supplement rmd (with corresponding figures)
- Manuscript figures rmd
