---
title: "\\helveticaitalic{Supplementary Material}"
author: "Eva Koster et al."
output: 
  pdf_document:
    keep_tex: true
    number_sections: true
    highlight: default
documentclass: frontiers-suppmat
classoption: utf8
bibliography: immune-reconst.bib
biblio-style: Frontiers-Vancouver
header-includes: 
  - \usepackage{url,hyperref,lineno,microtype,float}
  - \usepackage[onehalfspacing]{setspace}
  - \onecolumn
  - \firstpage{1}
  - \raggedbottom
editor_options: 
  chunk_output_type: console
  markdown: 
    wrap: 72
---

```{r setup, include=FALSE}
# Knitr set-up
knitr::opts_knit$set(root.dir = here::here())
knitr::opts_chunk$set(
  echo = FALSE, 
  out.width = "100%", 
  warning = FALSE, 
  message = FALSE,
  dev = "cairo_pdf" # important, or else Roboto Cond. font won't export
)

# https://stackoverflow.com/questions/55009313/how-to-use-cairo-pngs-in-r-markdown

# Version control: https://github.com/hughjonesd/latexdiffr for latexdiffr
```

<!-- Load targets objects -->

```{r load-objs}
# Load libraries, set seed and theme
set.seed(8498456)
library(extrafont)
library(Manu)
library(data.table)
library(splines)
library(targets)
library(ggplot2)
library(JM)

# Global plots theme + settings
colrs <- Manu::get_pal("Hoiho")
confint_alpha <- 0.5
confint_col <- "lightgray"
global_font <- "Roboto Condensed"
log_axis_scales <- scale_y_continuous(
  breaks = log(c(0.1, 1, 5, 25, 100, 500, 1500)),
  labels = c(0.1, 1, 5, 25, 100, 500, 1500)
)

theme_set(
  theme_light(base_size = 14, base_family = global_font) +
    theme(
      strip.background = element_rect(fill = colrs[[2]], colour = "transparent"),
      strip.text = element_text(colour = 'white')
    )
)

tar_load(
  c(
    NMA_preDLI_datasets,
    dat_merged
  )
)
```

The present supplementary material is a \`Statistical Supplement' to the
main article, outlining the methods used and the motivation behind them.

# Overview

The main article seeked to investigate the impact of early low-dose
donor lymphocyte infusion (DLI) on the T-cell kinetics and the
association between these kinetics and the development of clinical
events. The data at hand are comprised of

-   Baseline patient characteristics.
-   Absolute T-cell counts measured routinely at discrete timepoints
    (not exactly the same times across patients) starting from time of
    allogeneic stem cell transplantation (alloSCT).
-   Time of standard prophylactic DLI, and time of early low dose DLI
    (if given).
-   Time to various clinical events occurring during follow-up, such as
    start of therapeutic immunosuppression for treatment
    Graft-versus-Host-Disease (GvHD), or relapse of the underlying
    disease.

Note that the data components outlined above are inextricably
interlinked. For example, a patient dying or dropping out of a study
clearly precludes any further T-cell measurements from being collected.
In other words, the T-cell counts that would have been measured in the
absence of death or drop-out are *missing*. The suspected reason(s)
behind this missingness is a key consideration behind the choice of
analysis approach: the occurrence of death could be related to observed
(e.g. baseline covariates, measured T-cell counts) or unobserved (e.g.
latent characteristics of the T-cell counts such as the 'true' value at
a point in time, free of measurement error) information
[@rouanetInterpretationMixedModels2019].

While death and drop-out are well-known drivers of cohort attrition, for
modelling purposes the longitudinal measurements may also be
\`artificially terminated' at the occurrence of some clinical event of
interest (e.g. GvHD in the present context) after which the patient may
in reality remain in follow-up afterwards. Although it may be clear why
we should do this when modelling the association between a longitudinal
marker and a time-to-event outcome (future measurements cannot help
predict past events), it is also an important consideration for the
trajectory of the marker itself. The trajectory changes so fundamentally
through the clinical events that analyzing subsequent measurements
together with those preceding the event is generally not meaningful.

<!-- Mention multi state is a way to make it make sense -->

## Modelling the impact of early low-dose DLI on T-cell kinetics

In the main article, we performed an intention-to-treat (ITT) analysis,
comparing baseline disease risk groups on the basis of which early
low-dose DLIs were scheduled. A first avenue for assessing the impact of
early low-dose DLI on T-cell kinetics would be simple descriptive
analysis: for example, selecting a number of timepoints or intervals and
thereafter visualizing the distribution and average of the T-cell counts
per ITT group. Figure S1 below shows a possible analysis for the CD4
counts.

```{r descrip, eval=TRUE, fig.cap = "Descriptive analysis, dividing the measurements into one-month intervals.", fig.pos="H",fig.height=4}
dat_long <- NMA_preDLI_datasets$long
dat_wide <- NMA_preDLI_datasets$wide

dat_long[, time_cat := cut(
  intSCT2_7, include.lowest = TRUE, breaks = seq(0, 6, length.out = 7)
)]#ggplot2::cut_number(intSCT2_7, 8)]
dat_long |> 
  ggplot(aes(time_cat, CD4_abs_log, fill = hirisk)) +
  geom_boxplot() +
  facet_wrap(~ ATG) +
  scale_fill_manual(
    "Risk group",
    labels = c("Non-high risk", "High risk"),
    values = c(colrs[[4]], colrs[[1]])
  ) +
  scale_y_continuous(
    breaks = log(c(0.1, 1, 5, 25, 100, 500, 1500)),
    labels = c(0.1, 1, 5, 25, 100, 500, 1500)
  ) +
  labs(
    x = "Time interval (months since alloSCT)",
    y = expression(paste("CD4 (x10"^"6","/l)"))
  ) +
  theme(legend.position = "top")
```

```{r linmod-flex, fig.cap = "Analysis using a linear model with restricted cubic splines.", fig.pos="H", fig.height=4}
mod_lin <- lm(
  formula = CD4_abs_log ~ ns(intSCT2_7, 3) * hirisk * ATG,
  data = dat_long
)
dat_long[, pred_lin := predict(mod_lin)]

dat_long |> 
  ggplot(aes(intSCT2_7, CD4_abs_log)) +
  geom_point(size = 1.5, alpha = 0.5, col = colrs[[1]]) + 
  #geom_line(aes(group = IDAA), size = 1, alpha = 0.1) +
  #geom_boxplot() +
  facet_wrap(~ ATG) +
  geom_line(
    aes(y = pred_lin, group = hirisk, col = hirisk, linetype = hirisk),
    linewidth = 2
  ) + 
  scale_y_continuous(
    breaks = log(c(0.1, 1, 5, 25, 100, 500, 1500)),
    labels = c(0.1, 1, 5, 25, 100, 500, 1500)
  ) +
  labs(
    x = "Time since alloSCT (months)",
    y = expression(paste("CD4 (x10"^"6","/l)"))
  ) +
  scale_colour_manual(
    "Risk group",
    labels = c("Non-high risk", "High risk"),
    values = c(colrs[[4]], colrs[[6]])
  ) +
  scale_linetype_discrete(
    "Risk group",
    labels = c("Non-high risk", "High risk")
  ) +
  theme(legend.position = "top")
```

The measurements in Figure S1 are binned into one-month intervals, and are further split by risk group (non-high versus high risk) and donor type (related donor, RD, versus unrelated donor, UD). To simplify the visualisations in this section, the CMV patient-donor match is ignored.

Since conclusions based on a descriptive analysis may depend on the
chosen time intervals and number of measurements per interval, we may
instead choose to smooth across all individual measurements using a
(flexible) linear model (Figure S2). This however does not take into
account the correlation between different values of the same patient
over time. Figure S3 instead looks at the average and fitted individual
trajectories obtained with a linear mixed effects model.

```{r mixed-mod, fig.cap="Analysis using a linear mixed-effects model.", fig.pos="H", fig.height=4}
library(nlme)
mod_mixed <- lme(
  fixed = CD4_abs_log ~ ns(intSCT2_7, 3) * hirisk * ATG,
  data = dat_long, 
  random = ~ 0 + ns(intSCT2_7, 3) | IDAA,
  control = lmeControl(opt = "optim", msMaxIter = 200)
)

dat_long[, ':=' (
  pred_marg = predict(mod_mixed, level = 0L),
  pred_indiv = predict(mod_mixed, level = 1L)
)]

dat_long |> 
  ggplot(aes(intSCT2_7, pred_indiv)) +
  geom_line(aes(group = IDAA), size = 1, alpha = 0.5, col = colrs[[1]]) +
  #geom_boxplot() +
  facet_wrap(~ ATG) +
  geom_line(
    aes(y = pred_marg, group = hirisk, col = hirisk, linetype = hirisk),
    linewidth = 2
  ) + 
  scale_y_continuous(
    breaks = log(c(0.1, 1, 5, 25, 100, 500, 1500)),
    labels = c(0.1, 1, 5, 25, 100, 500, 1500)
  ) +
  labs(
    x = "Time since alloSCT (months)",
    y = expression(paste("CD4 (x10"^"6","/l)"))
  ) +
  scale_colour_manual(
    "Risk group",
    labels = c("Non-high risk", "High risk"),
    values = c(colrs[[4]], colrs[[6]])
  ) +
  scale_linetype_discrete(
    "Risk group",
    labels = c("Non-high risk", "High risk")
  ) +
  theme(legend.position = "top")
```

Note that the mixed model based analysis relies on the assumption that
the termination of the measurements can be explained only by observed
characteristics (i.e. baseline covariates or measured T-cell counts).

<!-- Forest plot of coefficients? -->

## Association between T-cell kinetics and clinical events

To assess the association between between the T-cell kinetics and the (competing) events described in the main article, we could use extended cause-specific Cox proportional hazards models. The T-cell measurements are then used as a time-varying predictor. This approach ignores any measurement error, and assumes that the value of the time-varying predictor stays *constant* between measurement times. As a result, the true association may be underestimated, and the estimate will have larger confidence intervals compared to those obtained using a joint model (described in the following section). A recent tutorial comparing the joint model to the extended Cox model is found in the work of @baartJointModelingLongitudinal2021. 

## Bringing together the two: a joint model

@rizopoulosJointModelsLongitudinal2012 provides a comprehensive overview shared-parameter joint models, and their extensions. A more compact overview can also be found in the work of [@papageorgiouOverviewJointModeling2019]. As described in the main article, joint models consist of two components: a longitudinal submodel (generally a mixed effects regression model), and a time-to-event submodel (a Cox-like proportional hazards model). Importantly, the random effects (used to capture the variability in subject-specific T-cell count trajectories) are *shared* between the models. This implies we can a) assess the impact of the T-cell trajectories on the hazard of the clinical events, and b) the estimated trajectories can themselves be corrected for bias related to the measurements being terminated by the occurrence of the clinical events. Furthermore, in contract to the time-varying Cox model, measurement error is correctly accounted for in joint models. The joint models used in the main article are described in detail below.

<!-- Comparison also of results with JM vs TD cox -->

```{r test-swimmer,fig.cap="Swimmer plot (to do: label DLI)", fig.pos="htbp", fig.dim=c(6, 8), eval=FALSE}
# Only works with out.width for now; place at end / beginning sections
# See also https://stackoverflow.com/questions/42486617/knitr-ignoring-fig-pos
# https://github.com/rstudio/rmarkdown/issues/1345

#dat_merged[TCD2 %in% c("NMA RD: ALT", "UD: ALT + ATG")]

# Prep endpoints 
dat_evs <- melt.data.table(
  data = dat_wide,
  id.vars = "IDAA",
  measure.vars = c("uDLI", "endpoint7"),
  variable.name = "event_id",
  value.name = "intSCT2_7"
)
#dat_evs[intSCT2_7 > 6]

# Actually better to do this with raw data

dat_long |> 
  ggplot(aes(x = intSCT2_7, y = reorder(IDAA, endpoint7))) +
  geom_line(aes(group = IDAA), linetype = "dotted", alpha = 0.5) +
  geom_point(
  ) +
  facet_grid(
    endpoint7_s ~ ., 
    #labeller = labeller(endpoint5_s = event_labs),
    scales = "free_y",
    space = "free"
  ) +
  labs(x = "Time since alloHCT (months)", y = "Patient ID") + 
  theme_light(base_size = 14) +
  theme(
    legend.position = "none",
    axis.text.y = element_blank(), 
    panel.grid.major.y = element_blank()
  ) 
```

# Joint Model I

## Model formulation

The (log) T-cell counts (either CD3, CD4 or CD8) for the $i^{\text{th}}$
patient at timepoints $t_{ij} \ (j = 1,\ldots,n_i)$ are denoted by
$y_i(t)$. For the longitudinal submodel, we postulated a linear
mixed-effects model as \begin{align*}
  y_i(t) &= m_i(t) + \epsilon_i(t), \\
         &= \beta_0 + \sum_{q=1}^3(\beta_q + b_{iq})B_q(t)  + \sum_{q=1}^3 \beta_{q + 3}\{B_q(t) \times \text{Risk}_i\} + \sum_{q=1}^3 \beta_{q + 6}\{B_q(t) \times \text{Donor}_i\} \\
  &\qquad + \sum_{q=1}^3 \beta_{q + 9}\{B_q(t) \times \text{Risk}_i \times \text{Donor}_i\} + \beta_{13}\text{CMV}_i + \beta_{14}\text{Risk}_i + \beta_{15}\text{Donor}_i \\
  &\qquad + \beta_{16} \{\text{Risk}_i \times \text{Donor}_i\} + \epsilon_i(t).
\end{align*} $\text{Risk}_i$, $\text{Donor}_i$ and $\text{CMV}_i$
respectively represent the dummy variables for baseline disease risk
(the intention-to-treat variable, high-risk compared to non-high risk),
donor type (unrelated compared to related donor) and patient/donor
Cytomegalovirus (CMV) serostatus at baseline (any one of patient or
donor positive, compared to patient and donor both negative).

Time since alloSCT was modelled flexibly assuming restricted (natural)
cubic splines with two internal knots placed at the 33.3% and 66.7%
percentiles of the measurement times. This is represented above by
$B_q(t)$, corresponding to the $q^{\text{th}}$ basis function of the
spline. The fixed effects part of the model posits a three-way
interaction between time, donor type and baseline disease risk, as well
as a main effect of patient/donor CMV status. The three-way interaction
was constructed to a) capture the slower expected average trajectory of
patients with an unrelated donor, due to the use of anti-thymocyte
globulin (ATG) in this group; and b) to test for a difference in average
trajectories between baseline disease risk groups.

In terms of random effects, this models assumes random slopes $b_{iq}$
(one for each basis function), and a fixed intercept. This fixed
intercept was justified given that this cohort underwent T-cell depleted
(TCD) alloSCT, and most patients were therefore expected to start
follow-up with T-cell counts around zero. The random slopes were assumed
to be normally distributed with mean zero, with unstructured covariance
matrix $D$. Furthermore, $\epsilon_i(t)$ is the patient specific error
term.

<!-- Code here with long format and nlme code? -->

The time-to-event submodel was composed of multiple cause-specific
proportional hazards models as \begin{align*}
  h_{1i}(t) &= h_{10}(t) \exp\big\{\gamma_{11}\text{Donor}_i + \gamma_{12}\text{Risk}_i + \alpha_1 m_i(t) \big\}, \\ 
  h_{2i}(t) &= h_{20}(t) \exp\big\{\gamma_{21}\text{Donor}_i + \gamma_{22}\text{Risk}_i + \alpha_2 m_i(t) \big\}, \\
  h_{3i}(t) &= h_{30}(t) \exp\big\{\gamma_{31}\text{Donor}_i + \alpha_3 m_i(t) \big\},
\end{align*} where the $h_{ki}(t)$ for $k \in \{1, 2, 3\}$ respectively
represent the cause-specific hazards of GvHD, relapse and other
failures. The cause-specific baseline hazards $h_{k0}(t)$ were
approximated on the log scale using cubic B-splines with three internal
knots. The above corresponds to the 'current value' parametrization of
the joint model, where the $\exp(\alpha_k)$ would represent the hazard
ratio (for cause $k$) when comparing two patients (with same covariates)
whose 'true' underlying (log) T-cell values at a particular timepoint
$m_i(t)$ differ by one. The $\gamma_{kp}$ coefficients are interpreted
analogously to main effects in standard cause-specific Cox proportional
hazards models.

In addition to the current value parametrization, we also ran the models
assuming a time-dependent slopes association structure as
$\alpha_{k1}m_i(t) + \alpha_{k2}\{d mi(t)/dt\}$.

## Diagnostics

```{r resid-preDLI}
purrr::map_dfr(
  list(
    "CD4" = tar_read(preDLI_JM_value_corr_CD4),
    "CD8" = tar_read(preDLI_JM_value_corr_CD8),
    "CD3" = tar_read(preDLI_JM_value_corr_CD3)
  ),
  .f = ~ {
    rbind.data.frame(
      cbind.data.frame(
        "resid" = residuals(.x, process = "Longitudinal", type = "stand-Subject"),
        "fitted" = fitted(.x, process = "Longitudinal", type = "Subject"),
        "type" = "Subject-specific"
      ), 
      cbind.data.frame(
        "resid" = residuals(.x, process = "Longitudinal", type = "stand-Marginal"),
        "fitted" = fitted(.x, process = "Longitudinal", type = "Marginal"),
        "type" = "Average"
      )
    )
  },
  .id = "cell"
) |> 
  ggplot(aes(fitted, resid)) + #exp(fitted)
  geom_point(alpha = 0.5) +
  geom_smooth(method = "loess", formula = y ~ x) +
  geom_hline(yintercept = 0L, linetype = "dashed") +
  facet_grid(cell ~ type, scales = "fixed") +
  labs(x = "Fitted values", y = "Standardized Residuals")
```

# Joint Model II

## Model formulation

For model II, the time scale was no longer from alloSCT, but instead
from time of early low-dose DLI. Therefore, this model was only run
among the subset that *did* in fact receive an early low-dose DLI before
the occurrence of other competing events. The longitudinal submodel was
again a linear mixed-effect model, given by \begin{align*}
  y_i(t) &= m_i(t) + \epsilon_i(t), \\
         &= (\beta_0 + b_{i0}) + \sum_{q=1}^2(\beta_q + b_{iq})B_q(t)  + \sum_{q=1}^2 \beta_{q + 2}\{B_q(t) \times \text{Donor}_i\} + \beta_{5}\text{CMV}_i + \epsilon_i(t).
\end{align*} Time was again modelled with restricted cubic splines, but
in contrast to model I, we used a single internal knot - due to smaller
sample size and fewer measurements per person. For consistency with
model I, this average trajectory was allowed to differ per donor type
(two-way interaction). In this model, disease risk at baseline was
redundant as we ran the model among those having actually received an
early low-dose DLI. A fixed effect for patient/donor CMV serostatus was
also added to the model. This model comprised both random intercepts
$b_{i0}$ and random slopes $b_{iq}$, assumed to follow normal
distributions with mean zero and unstructured covariance matrix.

Due to a limited number of events, relapse and other failure were merged
into a composite endpoint. The time-to-event submodel was therefore
specified as \begin{align*}
  h_{1i}(t) &= h_{10}(t) \exp\big\{\gamma_{11}\text{Donor}_i + \alpha_1 m_i(t) \big\}, \\ 
  h_{2i}(t) &= h_{20}(t) \exp\big\{\alpha_2 m_i(t) \big\},
\end{align*} where the $h_{ki}(t)$ for $k \in \{1, 2\}$ respectively
represent the cause-specific hazards of GvHD and the composite of
relapse and other failures. The cause-specific baseline hazards
$h_{k0}(t)$ were approximated on the log scale using cubic B-splines
with two internal knots. In this joint model, only the current value
parametrization was explored.

<!-- Debate whether to use pseudo code if github there anyway? -->

## Diagnostics

```{r resid-postDLI}
purrr::map_dfr(
  list(
    "CD4" = tar_read(postDLI_JM_corr_CD4),
    "CD8" = tar_read(postDLI_JM_corr_CD8),
    "CD3" = tar_read(postDLI_JM_corr_CD3)
  ),
  .f = ~ {
    rbind.data.frame(
      cbind.data.frame(
        "resid" = residuals(.x, process = "Longitudinal", type = "stand-Subject"),
        "fitted" = fitted(.x, process = "Longitudinal", type = "Subject"),
        "type" = "Subject-specific"
      ), 
      cbind.data.frame(
        "resid" = residuals(.x, process = "Longitudinal", type = "stand-Marginal"),
        "fitted" = fitted(.x, process = "Longitudinal", type = "Marginal"),
        "type" = "Average"
      )
    )
  },
  .id = "cell"
) |> 
  ggplot(aes(fitted, resid)) + #exp(fitted)
  geom_point(alpha = 0.5) +
  geom_smooth(method = "loess", formula = y ~ x) +
  geom_hline(yintercept = 0L, linetype = "dashed") +
  facet_grid(cell ~ type, scales = "free_x") +
  labs(x = "Fitted values", y = "Standardized Residuals")
```

# Other considerations

## Limitations and possible extensions

In the main article, separate joint models were run for each T-cell subset. Ideally, we would like to model these subset *jointly*, using a multivariate (i.e. allowing multiple correlated longitudinal outcomes) joint model. First, this would allow to have a single set of estimates for the main effects of donor type and baseline risk group, adjusted for all T-cell subsets. Currently, the estimates of these main effects (visualized in the forest plots from Figures 5 and 6 of the main article) differ simply because they are each adjust for a different T-cell subset in turn. Second, such a multivariate model would allow dynamic prediction of the competing events conditional on the current values (and slopes) of all T-cell subsets. Nevertheless, such models are generally computationally intensive, and likely require a much larger sample size.


Something about trajectories interpretation as immortal cohort or for a
living patient with random effects equal to zero
[@rouanetHowSelectionTime2022].

Something about possibility of modelling the intermediate DLI explicity
in the trajectories as
[@papageorgiouIndividualizedDynamicPrediction2019], or instead use
multi-state submodel [Ferrer reference].

Also possibly reference Cuer paper on competing risks joint models.

## Software choices

While the `{JM}` R package [@rizopoulosJMPackageJoint2010] was used in the current work, many different R packages implement shared parameter joint models. The most notable others are `{JMbayes}` [@rizopoulosPackageJMbayesFitting2016], `{JMbayes2}` [@rizopoulosJMbayes2ExtendedJoint2022], `{joineR}` [@philipsonJoineRJointModelling2018,@williamsonJointModellingLongitudinal2008] and `{joineRML}` [@hickeyJoineRMLJointModel2018]. Each of these differ either based on the estimation framework and methods (e.g. Bayesian, frequentist), or on the specific parametrizations used for the joint model (e.g. shared random effects or latent).
@hickeyComparisonJointModels2018 provide an overview of joint models for competing risks.

# References
