---
title: "\\helveticaitalic{Supplementary Material}"
author: "Eva Koster et al."
output: 
  pdf_document:
    keep_tex: true
    number_sections: true
    highlight: default
documentclass: frontiers-suppmat
classoption: utf8
bibliography: immune-reconst.bib
biblio-style: Frontiers-Vancouver
header-includes: 
  - \usepackage{url,hyperref,lineno,microtype,float}
  - \usepackage[onehalfspacing]{setspace}
  - \onecolumn
  - \firstpage{1}
---

```{r setup, include=FALSE}
# Knitr set-up
knitr::opts_knit$set(root.dir = here::here())
knitr::opts_chunk$set(
  echo = FALSE, 
  out.width = "100%", 
  warning = FALSE, 
  message = FALSE
)

# Version control: https://github.com/hughjonesd/latexdiffr for latexdiffr

# Load libraries, set seed and theme
set.seed(8498456)
library(data.table)
library(targets)
library(ggplot2)

theme_set(theme_bw(base_size = 14))
```

<!-- Load targets objects -->

```{r load-objs}
tar_load(
  c(
    NMA_preDLI_datasets,
    dat_merged
  )
)
```

The present supplementary material is a `Statistical Supplement' to the main article, outlining the methods used and the motivation behind them.

# Research questions and data overview

The two primary research questions addressed in the main article are

1. What is the effect of an early low dose donor lymphocyte infusion (DLI) on the cell count kinetics within the T-cell compartment during the first 6 months after allogeneic stem cell transplantation (alloSCT)?

2. Are the T-cell kinetics from time of early low dose DLI associated with the risk of Graft-versus-Host-Disease (GvHD)?

To answer both questions, the data at hand are comprised of

- Baseline patient characteristics.
- T-cell absolute counts measured routinely at discrete timepoints starting from time of alloSCT.
- Time of standard prophylactic DLI, and time of early low dose DLI (if given).
- Time to various clinical events occuring during follow-up, such as GvHD or relapse of the underlying disease.

Note that the data components outlined above are inextricably interlinked. For example, a patient dying or dropping out of a study clearly precludes any further T-cell measurements from being collected. In other words, the T-cell counts that would have been measured in the absence of death or drop-out are *missing*. The suspected reason(s) behind this missingness is a key consideration behind in the choice of analysis aimed at answering the research questions: the occurrence of death could be related to observed (e.g. baseline covariates, measured T-cell counts) or unobserved (e.g. latent characteristics of the T-cell counts, such as speed of increase/decrease at a point in time) information [@rouanetInterpretationMixedModels2019].

While death and drop-out are well-known drivers of cohort attrition, longitudinal measurements may also be `artificially terminated' at the occurrence of some intermediate event after which the patient remains in follow-up. This is justified when for example in the case of GVHD: no use of using measurements after to predict event in the past. It can also be justified when an event is expected to modify the trajectories?

```{r test-swimmer,fig.cap="Swimmer plot (to do: label DLI)", fig.pos="htbp", fig.dim=c(6, 8)}
# Only works with out.width for now; place at end / beginning sections
# See also https://stackoverflow.com/questions/42486617/knitr-ignoring-fig-pos
# https://github.com/rstudio/rmarkdown/issues/1345
dat_long <- NMA_preDLI_datasets$long
dat_wide <- NMA_preDLI_datasets$wide
#dat_merged[TCD2 %in% c("NMA RD: ALT", "UD: ALT + ATG")]

# Prep endpoints 
dat_evs <- melt.data.table(
  data = dat_wide,
  id.vars = "IDAA",
  measure.vars = c("uDLI", "endpoint7"),
  variable.name = "event_id",
  value.name = "intSCT2_7"
)
#dat_evs[intSCT2_7 > 6]

# Actually better to do this with raw data

dat_long |> 
  ggplot(aes(x = intSCT2_7, y = reorder(IDAA, endpoint7))) +
  geom_line(aes(group = IDAA), linetype = "dotted", alpha = 0.5) +
  geom_point(
  ) +
  facet_grid(
    endpoint7_s ~ ., 
    #labeller = labeller(endpoint5_s = event_labs),
    scales = "free_y",
    space = "free"
  ) +
  labs(x = "Time since alloHCT (months)", y = "Patient ID") + 
  theme_light(base_size = 14) +
  theme(
    legend.position = "none",
    axis.text.y = element_blank(), 
    panel.grid.major.y = element_blank()
  ) 
```

# Effect of early low dose DLI on cell count kinetics

## Simple approaches

<Naive approaches, descriptives, dividing based on endpoint
Linear models, mixed models, Rouanet paper, defining a population
Choices to make in joint model, additional complexity with each extra competing risk
Inadequate sample size, baseline hazard complexity, optimisation complexity
Needing a survival model even though not necessarily interested in survival> 

[Write some stuff here, organise placeholders for plots, divide stuff up later, find references in Zotero]

## Pre-DLI model

Example references @rizopoulosJointModelsLongitudinal2012

$D = \{0, 1, 2, 3\}$ where 0 = *censored*, 1 = *GvHD*, 2 = *Relapse* and 3 = *Other failures*.

\begin{align*}
  h_{1i}(t) &= h_{10}(t) \exp\big\{\gamma_{11}\text{UD}_i + \gamma_{12}\text{ITT}_i + \alpha_1 m_i(t) \big\} \\ 
  h_{2i}(t) &= h_{20}(t) \exp\big\{\gamma_{21}\text{UD}_i + \gamma_{22}\text{ITT}_i + \alpha_2 m_i(t) \big\} \\
  h_{3i}(t) &= h_{30}(t) \exp\big\{\gamma_{31}\text{UD}_i + \alpha_3 m_i(t) \big\}
\end{align*}

For longitudinal we assume $y_i(t) = m_i(t) + \epsilon_i(t)$, where


\begin{align*}
  m_i(t) &= \beta_0 + \sum_{q=1}^3(\beta_q + b_{iq})B_n(t,\lambda_q)  + \sum_{q=1}^3 \beta_{q + 3}\{B_n(t,\lambda_q) \times \text{ITT}_i\} + \sum_{q=1}^3 \beta_{q + 6}\{B_n(t,\lambda_q) \times \text{UD}_i\} + \\
  & \sum_{q=1}^3 \beta_{q + 9}\{B_n(t,\lambda_q) \times \text{ITT}_i \times \text{UD}_i\} + \beta_{13}\text{CMV}_i + \beta_{14}\text{ITT}_i + \beta_{15}\text{UD}_i + \beta_{16} \{\text{ITT}_i \times \text{UD}_i\}
\end{align*}

# Association between T-cell kinetics and GVHD after early low dose DLI

## Post-DLI model

$D = \{0, 1, 2\}$ where 0 = *censored*, 1 = *GvHD*, 2 = *Relapse or other failures*.

\begin{align*}
  h_{1i}(t) &= h_{10}(t) \exp\big\{\gamma_{11}\text{EarlyDLI}_i + \alpha_1 m_i(t) \big\} \\ 
  h_{2i}(t) &= h_{20}(t) \exp\big\{\alpha_2 m_i(t) \big\}
\end{align*}

And longitudinal 

\begin{align*}
  m_i(t) &= (\beta_0 + b_{i0}) + \sum_{q=1}^2(\beta_q + b_{iq})B_n(t,\lambda_q)  + \sum_{q=1}^2 \beta_{q + 2}\{B_n(t,\lambda_q) \times \text{EarlyDLI}_i\} + \beta_{15}\text{UD}_i 
\end{align*}

## Subsection

...

## Subsubting

Illustration of slope

Test code snippet

```{r test-snip, echo=TRUE, eval=FALSE}
jointModel(
  lmeObject = preDLI_long_corr_CD3,
  survObject = preDLI_cox,
  CompRisk = TRUE,
  method = "spline-PH-aGH",
  timeVar = "intSCT2_7",
  parameterization = "value",
  interFact = list("value" = ~ strata(trans) - 1),
  iter.qN = 1000,
  lng.in.kn = preDLI_basehaz_knots,
  numeriDeriv = "cd",
  eps.Hes = 1e-04
)
```

# General considerations

## Software choices

JM, JMbayes, JMbayes2, joineR, joineRML.. hickeyComparisonJointModels2018

## Optimisation in JM

## Baseline hazards

Number of knots, lalala

# References
