---
title: "\\helveticaitalic{Supplementary Material}"
author: "Eva Koster et al."
output: 
  pdf_document:
    keep_tex: true
    number_sections: true
    highlight: default
documentclass: frontiers-suppmat
classoption: utf8
bibliography: immune-reconst.bib
biblio-style: Frontiers-Vancouver
header-includes: 
  - \usepackage{url,hyperref,lineno,microtype,float}
  - \usepackage[onehalfspacing]{setspace}
  - \onecolumn
  - \firstpage{1}
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
# Knitr set-up
knitr::opts_knit$set(root.dir = here::here())
knitr::opts_chunk$set(
  echo = FALSE, 
  out.width = "100%", 
  warning = FALSE, 
  message = FALSE
)

# Version control: https://github.com/hughjonesd/latexdiffr for latexdiffr

# Load libraries, set seed and theme
set.seed(8498456)
library(data.table)
library(targets)
library(ggplot2)
library(JM)

theme_set(theme_bw(base_size = 14))
```

<!-- Load targets objects -->

```{r load-objs}
tar_load(
  c(
    NMA_preDLI_datasets,
    dat_merged
  )
)
```

The present supplementary material is a `Statistical Supplement' to the main article, outlining the methods used and the motivation behind them.

# Overview

The main article seeked to investigate the impact of early low-dose donor lymphocyte infusion (DLI) on the T-cell kinetics and the association between these kinetics and the development of clinical events. The data at hand are comprised of

- Baseline patient characteristics.
- Absolute T-cell counts measured routinely at discrete timepoints (not exactly the same times across patients) starting from time of allogeneic stem cell transplantation (alloSCT).
- Time of standard prophylactic DLI, and time of early low dose DLI (if given).
- Time to various clinical events occurring during follow-up, such as Graft-versus-
Host-Disease (GvHD) or relapse of the underlying disease.

Note that the data components outlined above are inextricably interlinked. For example, a patient dying or dropping out of a study clearly precludes any further T-cell measurements from being collected. In other words, the T-cell counts that would have been measured in the absence of death or drop-out are *missing*. The suspected reason(s) behind this missingness is a key consideration behind the choice of analysis approach: the occurrence of death could be related to observed (e.g. baseline covariates, measured T-cell counts) or unobserved (e.g. latent characteristics of the T-cell counts, such as rate of increase at a point in time) information [@rouanetInterpretationMixedModels2019].

While death and drop-out are well-known drivers of cohort attrition, for modelling purposes the longitudinal measurements may also be `artificially terminated' at the occurrence of some clinical event of interest (e.g. GvHD in the present context) after which the patient may in reality remain in follow-up afterwards. While it may be clear why we should do this when modelling the association between a longitudinal marker and a time-to-event outcome (future measurements cannot help predict past events), it is also an important consideration for the trajectory of the marker itself: the population of interest evolves over time, and the clinical may influence the trajectories.

## Modelling the impact of early low-dose DLI on T-cell kinetics

Descriptives, linear models, mixed models, joint models

Reference on intermediate event trajectory paper Greg

Joint model population currently alive estimand/random effect set to zero

## Association between T-cell kinetics and clinical events

Time-dependent Cox suboptimal, added complexity with each comp risk (Cuer paper?)

## Bringing together the two: a joint model

Cite Rizopoulos, and Greg JMs overview paper

```{r test-swimmer,fig.cap="Swimmer plot (to do: label DLI)", fig.pos="htbp", fig.dim=c(6, 8), eval=FALSE}
# Only works with out.width for now; place at end / beginning sections
# See also https://stackoverflow.com/questions/42486617/knitr-ignoring-fig-pos
# https://github.com/rstudio/rmarkdown/issues/1345
dat_long <- NMA_preDLI_datasets$long
dat_wide <- NMA_preDLI_datasets$wide
#dat_merged[TCD2 %in% c("NMA RD: ALT", "UD: ALT + ATG")]

# Prep endpoints 
dat_evs <- melt.data.table(
  data = dat_wide,
  id.vars = "IDAA",
  measure.vars = c("uDLI", "endpoint7"),
  variable.name = "event_id",
  value.name = "intSCT2_7"
)
#dat_evs[intSCT2_7 > 6]

# Actually better to do this with raw data

dat_long |> 
  ggplot(aes(x = intSCT2_7, y = reorder(IDAA, endpoint7))) +
  geom_line(aes(group = IDAA), linetype = "dotted", alpha = 0.5) +
  geom_point(
  ) +
  facet_grid(
    endpoint7_s ~ ., 
    #labeller = labeller(endpoint5_s = event_labs),
    scales = "free_y",
    space = "free"
  ) +
  labs(x = "Time since alloHCT (months)", y = "Patient ID") + 
  theme_light(base_size = 14) +
  theme(
    legend.position = "none",
    axis.text.y = element_blank(), 
    panel.grid.major.y = element_blank()
  ) 
```

# Joint Model I

## Model formulation

The (log) T-cell counts (either CD3, CD4 or CD8) for the $i^{\text{th}}$ patient at timepoints $t_{ij} \ (j = 1,\ldots,n_i)$ are denoted by $y_i(t)$. For the longitudinal submodel, we postulated a linear mixed-effects model as 
\begin{align*}
  y_i(t) &= m_i(t) + \epsilon_i(t), \\
         &= \beta_0 + \sum_{q=1}^3(\beta_q + b_{iq})B_q(t)  + \sum_{q=1}^3 \beta_{q + 3}\{B_q(t) \times \text{Risk}_i\} + \sum_{q=1}^3 \beta_{q + 6}\{B_q(t) \times \text{Donor}_i\} \\
  &\qquad + \sum_{q=1}^3 \beta_{q + 9}\{B_q(t) \times \text{Risk}_i \times \text{Donor}_i\} + \beta_{13}\text{CMV}_i + \beta_{14}\text{Risk}_i + \beta_{15}\text{Donor}_i \\
  &\qquad + \beta_{16} \{\text{Risk}_i \times \text{Donor}_i\} + \epsilon_i(t).
\end{align*}
$\text{Risk}_i$, $\text{Donor}_i$ and $\text{CMV}_i$ respectively represent the dummy variables for baseline disease risk (the intention-to-treat variable, high-risk compared to non-high risk), donor type (unrelated compared to related donor) and patient/donor Cytomegalovirus (CMV) serostatus at baseline (any one of patient or donor positive, compared to patient and donor both negative).

Time since alloSCT was modelled flexibly assuming restricted (natural) cubic splines with two internal knots placed at the 33.3\% and 66.7% percentiles of the measurement times. This is represented above by $B_q(t)$, corresponding to the $q^{\text{th}}$ basis function of the spline. The fixed effects part of the model posits a three-way interaction between time, donor type and baseline disease risk, as well as a main effect of patient/donor CMV status. The three-way interaction was constructed to a) capture the slower expected average trajectory of patients with an unrelated donor, due to the use of anti-thymocyte globulin (ATG) in this group; and b) to test for a difference in average trajectories between baseline disease risk groups. 

In terms of random effects, this models assumes random slopes $b_{iq}$ (one for each basis function), and a fixed intercept. This fixed intercept was justified given that this cohort underwent T-cell depleted (TCD) alloSCT, and most patients were therefore expected to start follow-up with T-cell counts around zero. The random slopes were assumed to be normally distributed with mean zero, with unstructured covariance matrix $D$. Furthermore, $\epsilon_i(t)$ is the patient specific error term.

<!-- Code here with long format and nlme code? -->

The time-to-event submodel was composed of multiple cause-specific proportional hazards models as 
\begin{align*}
  h_{1i}(t) &= h_{10}(t) \exp\big\{\gamma_{11}\text{Donor}_i + \gamma_{12}\text{Risk}_i + \alpha_1 m_i(t) \big\}, \\ 
  h_{2i}(t) &= h_{20}(t) \exp\big\{\gamma_{21}\text{Donor}_i + \gamma_{22}\text{Risk}_i + \alpha_2 m_i(t) \big\}, \\
  h_{3i}(t) &= h_{30}(t) \exp\big\{\gamma_{31}\text{Donor}_i + \alpha_3 m_i(t) \big\},
\end{align*}
where the $h_{ki}(t)$ for $k \in \{1, 2, 3\}$ respectively represent the cause-specific hazards of GvHD, relapse and other failures. The cause-specific baseline hazards $h_{k0}(t)$ were approximated on the log scale using cubic B-splines with three internal knots. The above corresponds to the 'current value' parametrization of the joint model, where the $\exp(\alpha_k)$ would represent the hazard ratio (for cause $k$) when comparing two patients (with same covariates) whose 'true' underlying (log) T-cell values at a particular timepoint $m_i(t)$ differ by one. The $\gamma_{kp}$ coefficients are interpreted analogously to main effects in standard cause-specific Cox proportional hazards models.

In addition to the current value parametrization, we als ran the models assuming a time-dependent slopes association structure as $\alpha_{k1}m_i(t) + \alpha_{k2}\{d mi(t)/dt\}$.

## Diagnostics

```{r resid-preDLI}
purrr::map_dfr(
  list(
    "CD4" = tar_read(preDLI_JM_value_corr_CD4),
    "CD8" = tar_read(preDLI_JM_value_corr_CD8),
    "CD3" = tar_read(preDLI_JM_value_corr_CD3)
  ),
  .f = ~ {
    rbind.data.frame(
      cbind.data.frame(
        "resid" = residuals(.x, process = "Longitudinal", type = "stand-Subject"),
        "fitted" = fitted(.x, process = "Longitudinal", type = "Subject"),
        "type" = "Subject-specific"
      ), 
      cbind.data.frame(
        "resid" = residuals(.x, process = "Longitudinal", type = "stand-Marginal"),
        "fitted" = fitted(.x, process = "Longitudinal", type = "Marginal"),
        "type" = "Average"
      )
    )
  },
  .id = "cell"
) |> 
  ggplot(aes(fitted, resid)) + #exp(fitted)
  geom_point(alpha = 0.5) +
  geom_smooth(method = "loess", formula = y ~ x) +
  geom_hline(yintercept = 0L, linetype = "dashed") +
  facet_grid(cell ~ type, scales = "fixed") +
  labs(x = "Fitted values", y = "Standardized Residuals")
```

# Joint Model II

## Model formulation

For model II, the time scale was no longer from alloSCT, but instead from time of early low-dose DLI. Therefore, this model was only run among the subset that *did* in fact receive an early low-dose DLI before the occurrence of other competing events. The longitudinal submodel was again a linear mixed-effect model, given by
\begin{align*}
  y_i(t) &= m_i(t) + \epsilon_i(t), \\
         &= (\beta_0 + b_{i0}) + \sum_{q=1}^2(\beta_q + b_{iq})B_q(t)  + \sum_{q=1}^2 \beta_{q + 2}\{B_q(t) \times \text{Donor}_i\} + \beta_{5}\text{CMV}_i + \epsilon_i(t).
\end{align*}
Time was again modelled with restricted cubic splines, but in contrast to model I, we used a single internal knot - due to smaller sample size and fewer measurements per person. For consistency with model I, this average trajectory was allowed to differ per donor type (two-way interaction). In this model, disease risk at baseline was redundant as we ran the model among those having actually received an early low-dose DLI. A fixed effect for patient/donor CMV serostatus was also added to the model. This model comprised both random intercepts $b_{i0}$ and random slopes $b_{iq}$, assumed to follow normal distributions with mean zero and unstructured covariance matrix.

Due to a limited number of events, relapse and other failure were merged into a composite endpoint. The time-to-event submodel was therefore specified as
\begin{align*}
  h_{1i}(t) &= h_{10}(t) \exp\big\{\gamma_{11}\text{Donor}_i + \alpha_1 m_i(t) \big\}, \\ 
  h_{2i}(t) &= h_{20}(t) \exp\big\{\alpha_2 m_i(t) \big\},
\end{align*}
where the $h_{ki}(t)$ for $k \in \{1, 2\}$ respectively represent the cause-specific hazards of GvHD and the composite of relapse and other failures. The cause-specific baseline hazards $h_{k0}(t)$ were approximated on the log scale using cubic B-splines with two internal knots. In this joint model, only the current value parametrization was explored.

<!-- Debate whether to use pseudo code if github there anyway? -->

## Diagnostics

```{r resid-postDLI}
purrr::map_dfr(
  list(
    "CD4" = tar_read(postDLI_JM_corr_CD4),
    "CD8" = tar_read(postDLI_JM_corr_CD8),
    "CD3" = tar_read(postDLI_JM_corr_CD3)
  ),
  .f = ~ {
    rbind.data.frame(
      cbind.data.frame(
        "resid" = residuals(.x, process = "Longitudinal", type = "stand-Subject"),
        "fitted" = fitted(.x, process = "Longitudinal", type = "Subject"),
        "type" = "Subject-specific"
      ), 
      cbind.data.frame(
        "resid" = residuals(.x, process = "Longitudinal", type = "stand-Marginal"),
        "fitted" = fitted(.x, process = "Longitudinal", type = "Marginal"),
        "type" = "Average"
      )
    )
  },
  .id = "cell"
) |> 
  ggplot(aes(fitted, resid)) + #exp(fitted)
  geom_point(alpha = 0.5) +
  geom_smooth(method = "loess", formula = y ~ x) +
  geom_hline(yintercept = 0L, linetype = "dashed") +
  facet_grid(cell ~ type, scales = "free_x") +
  labs(x = "Fitted values", y = "Standardized Residuals")
```

# General considerations

<!-- Example snippet-->

```{r test-snip, echo=FALSE, eval=FALSE}
jointModel(
  lmeObject = preDLI_long_corr_CD3,
  survObject = preDLI_cox,
  CompRisk = TRUE,
  method = "spline-PH-aGH",
  timeVar = "intSCT2_7",
  parameterization = "value",
  interFact = list("value" = ~ strata(trans) - 1),
  iter.qN = 1000,
  lng.in.kn = preDLI_basehaz_knots,
  numeriDeriv = "cd",
  eps.Hes = 1e-04
)
```

## Software choices

`{JM}`, `{JMbayes}`, `{JMbayes2}`, `{joineR}`, `{joineRML}`.. hickeyComparisonJointModels2018

## Optimisation in JM

(Probably skip)

# References
