---
title: "Immune reconstitution joint modelling"
author: "Ed Bonneville"
date: "`r Sys.setenv(LANG = 'en_US.UTF-8'); format(Sys.Date(), '%d %B %Y')`"
output: 
  html_document:
    df_print: kable
    toc: yes
    toc_float: 
      smooth_scroll: false
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_knit$set(root.dir = here::here())
knitr::opts_chunk$set(
  echo = FALSE, 
  out.width = "100%", 
  warning = FALSE, 
  message = FALSE
)

options(knitr.kable.NA = '', scipen=999)

# Load libraries
library(ggplot2)
library(data.table)
library(magrittr)
library(kableExtra)
library(ggrepel)

# Source in helpers
source("R/individual-cell-models.R")

# Prep ggplot2 labellers
atg_labs <- as_labeller(x = c("noATG" = "No ATG", "yesATG" = "ATG"))
cmv_labs <- as_labeller(
  x = c("Negative" = "Patient CMV: negative", "Positive" = "Patient CMV: positive")
)
event_labs <- as_labeller(
  x = c(
    "cens" = "Censored", 
    "cell_interv" = "Cellular Interv.",
    "REL" = "Relapse",
    "NRF_other" = "NRF: Other",
    "NRF_gvhd" = "NRF: GVHD"
  )
)
cell_labs <- as_labeller(
  x = c(
    "CD4_abs_log" = "CD4", 
    "CD8_abs_log" = "CD8",
    "NK_abs_log" = "NK"
  )
)
```


```{r data_read, echo=FALSE}
# Read-in merged (raw data)
tar_load(dat_merged)

# Get wide and long formats
datasets <- prepare_JM_data(
  merged_data = dat_merged, 
  cells = paste0(c("NK", "CD4", "CD8"), "_abs_log"), 
  admin_cens_time = 7
)

# Make a newdata to predict later
newdat <- expand.grid(
  "ATG" = levels(datasets$wide$ATG),
  "VCMVPAT_pre" = levels(datasets$wide$VCMVPAT_pre),
  "intSCT2_5" = seq(0.1, 6, by = 0.1)
)
```

# Descriptives

## Trajectories by competing event

```{r events_traj}
datasets$long %>% 
  melt.data.table(
    id.vars = c("IDAA", "intSCT2_5", "endpoint5_s"),
    variable.name = "cell_type", 
    value.name = "counts", 
    measure.vars = paste0(c("CD4", "CD8", "NK"), "_abs_log")
  ) %>% 
  ggplot(aes(x = intSCT2_5, y = counts)) +
  geom_line(aes(group = IDAA), alpha = 0.75, col = "gray") + 
  geom_smooth(method = "loess", formula = y ~ x, se = FALSE, col = "black") + 
  facet_grid(
    cell_type ~ endpoint5_s,
    labeller = labeller(cell_type = cell_labs, endpoint5_s = event_labs)
  ) + 
  scale_y_continuous(
    breaks = log(c(5, 25, 100, 500, 1500)),
    labels = c(5, 25, 100, 500, 1500)
  ) +
  labs(x = "Time since alloHCT (months)", y = "Cell counts") + 
  theme_bw() +
  theme(legend.position = "none") 
```

## Measurment schedule

```{r measurements}
# Try swimmer plot
datasets$long[, max_t := max(intSCT2_5), by = "IDAA"] 

datasets$long[, IDAA := reorder(IDAA, max_t)] %>% 
  ggplot(aes(x = intSCT2_5, y = IDAA)) + 
  geom_line(aes(group = IDAA), linetype = "dotted", alpha = 0.5) + 
  geom_point(aes(col = IDAA)) +
  facet_grid(
    endpoint5_s ~ ., 
    labeller = labeller(endpoint5_s = event_labs),
    scales = "free_y"
  ) +
  labs(x = "Time since alloHCT (months)", y = "IDAA") + 
  theme_bw() +
  theme(
    legend.position = "none",
    axis.text.y = element_blank()
  ) 
```

## DLI and trajectories

```{r dli_illust}
#... in CD8 comprisk file
dat_dli <- data.table::copy(dat_merged)
dat_dli <- dat_dli[!is.na(CD8_abs) & intSCT2_2 <= 24, .SD[.N > 1], by = "IDAA"]
dat_dli <- dat_dli[grepl(x = endpoint_specify, pattern = "DLI")] #DLI
dat_dli <- dat_dli[, post_DLI := factor(
  as.numeric(intSCT2_2 > endpoint2),
  levels = c(0, 1),
  labels = c("pre-DLI", "post-DLI")
)]


set.seed(86)
ref_pats <- sample(dat_dli$IDAA, size = 8, replace = FALSE)

dat_dli[IDAA %in% ref_pats] %>%
  ggplot(aes(intSCT2_5, CD8_abs_log)) +
  geom_point(
    aes(fill = post_DLI),
    size = 3.5, pch = 21,
    alpha = 0.8,
    col = "#359fda",
    #fill = colorspace::lighten("#359fda", amount = 0.3)
  ) +
  geom_path(alpha = 0.25, linetype = "dotted") +
  geom_vline(aes(xintercept = endpoint2), linetype = "dashed") +
  facet_wrap(~ IDAA, nrow = 2, scales = "free_x") +
  labs(x = "Time since alloHCT (months)", y = "CD8 cell counts") +
  scale_y_continuous(
    breaks = log(c(5, 25, 100, 500, 1500)),
    labels = c(5, 25, 100, 500, 1500)
  ) +
  theme_minimal() +
  guides(fill = guide_legend(title = "DLI"))
```

# Models

```{r load_mods}
tar_load(CD8_model_noInter)
tar_load(CD8_model_Inter)
tar_load(CD4_model_noInter)
tar_load(CD4_model_Inter)
tar_load(NK_model_noInter)
tar_load(NK_model_Inter)

# Put into lists
ls_mods_noInter <- list(
  "CD4" = CD4_model_noInter, 
  "CD8" = CD8_model_noInter,
  "NK" = NK_model_noInter
)

ls_mods_Inter <- list(
  "CD4" = CD4_model_Inter, 
  "CD8" = CD8_model_Inter,
  "NK" = NK_model_Inter
)

# Make the marginal predictions
preds_df_noInter <- data.table::rbindlist(
  lapply(ls_mods_noInter, function(mod) {
    JM::predict.jointModel(
      mod,
      newdata = newdat,
      type = "Marginal",
      idVar = "IDAA",
      returnData = TRUE,
      interval = "confidence"
    )
  }),
  idcol = "cell_type"
)

preds_df_Inter <- data.table::rbindlist(
  lapply(ls_mods_Inter, function(mod) {
    JM::predict.jointModel(
      mod,
      newdata = newdat,
      type = "Marginal",
      idVar = "IDAA",
      returnData = TRUE,
      interval = "confidence"
    )
  }),
  idcol = "cell_type"
)

# Prep ATG labels
by_vars <- c("cell_type", "ATG", "VCMVPAT_pre")
preds_df_noInter[intSCT2_5 == max(intSCT2_5), label := ifelse(
  ATG == "noATG", "ALT", "ALT + \n ATG"
), by = by_vars]
```

## CD8 joint model {.tabset}

### Specification

**Longitudinal sub-model:**

\begin{align*}
  \text{log(CD8)} \sim \text{ATG} \times \text{Spline(Time)} + \text{CMV (Patient)},
\end{align*}

with $\text{Spline(Time)}$ being a flexible, natural *cubic* spline. Patient-specific spline parameters (three random effects)

**Competing risks sub-model:**

\begin{align*}
  \text{(1) DLI} &\sim \text{SCT (pre-2010)} + \text{Disease Risk}, \\
  \text{(2) Relapse} &\sim \text{ATG} + \text{Disease Risk} + \text{log(CD8)}^*, \\
  \text{(3) Non-Relapse Failure: other} &\sim \text{ATG} + \text{log(CD8)}^*, \\
  \text{(4) Non-Relapse Failure: GVHD} &\sim \text{log(CD8)}^*, 
\end{align*}

with $\text{log(CD8)}^*$ = "true" *current value* of log(CD8) cell counts.

### Event submodel 

```{r cd8_nointer_table}
summarise_event_submodel(CD8_model_noInter) %>% 
  kableExtra::kbl(format = "html") %>%
  kable_paper("striped", full_width = F) %>%
  pack_rows("Cell. intervention", 1, 2) %>%
  pack_rows("Relapse", 3, 5) %>%
  pack_rows("NRF: Other", 6, 7) %>%
  pack_rows("NRF: GVHD", 8, 8)
```

### Predicted trajectories

```{r cd8_traject}
preds_df_noInter[cell_type == "CD8"] %>%
  ggplot(aes(x = intSCT2_5, y = pred, group = ATG, col = ATG)) +
  geom_ribbon(
    aes(ymin = low, ymax = upp),
    fill = "gray", 
    alpha = 0.5, 
    col = NA
  ) +
  geom_line(size = 1.5, alpha = 0.75) +
  #geom_text(aes(label = label), hjust = 0, na.rm = TRUE, fontface = "bold") + 
  geom_label_repel(
    aes(label = label), 
    fontface = "bold",
    na.rm = TRUE, 
    hjust = 0,
    direction = "y",
    nudge_x = 0.5
  ) + 
  facet_wrap(facets = ~ VCMVPAT_pre, labeller = labeller(VCMVPAT_pre = cmv_labs)) +
  labs(x = "Time since alloHCT (months)", y = "CD8 cell counts") +
  scale_y_continuous(
    breaks = log(c(5, 25, 100, 500, 1500)),
    labels = c(5, 25, 100, 500, 1500)
  ) +
  theme_minimal() +
  theme(legend.position = "none") +
  coord_cartesian(xlim = c(0, 7.5)) +
  scale_color_brewer(palette = "Paired")
```

## CD4 joint model {.tabset}

### Specification

**Longitudinal sub-model:**

\begin{align*}
  \text{log(CD4)} \sim \text{ATG} \times \text{Spline(Time)} + \text{CMV (Patient)},
\end{align*}

with $\text{Spline(Time)}$ being a flexible, natural *cubic* spline. Patient-specific spline parameters (three random effects)

**Competing risks sub-model:**

\begin{align*}
  \text{(1) DLI} &\sim \text{SCT (pre-2010)} + \text{Disease Risk}, \\
  \text{(2) Relapse} &\sim \text{ATG} + \text{Disease Risk} + \text{log(CD4)}^*, \\
  \text{(3) Non-Relapse Failure: other} &\sim \text{ATG} + \text{log(CD4)}^*, \\
  \text{(4) Non-Relapse Failure: GVHD} &\sim \text{log(CD4)}^*, 
\end{align*}

with $\text{log(CD4)}^*$ = "true" *current value* of log(CD4) cell counts.

### Event submodel 

```{r cd4_nointer_table}
summarise_event_submodel(CD4_model_noInter) %>% 
  kableExtra::kbl(format = "html") %>%
  kable_paper("striped", full_width = F) %>%
  pack_rows("Cell. intervention", 1, 2) %>%
  pack_rows("Relapse", 3, 5) %>%
  pack_rows("NRF: Other", 6, 7) %>%
  pack_rows("NRF: GVHD", 8, 8)
```

### Predicted trajectories

```{r cd4_traject}
preds_df_noInter[cell_type == "CD4"] %>%
  ggplot(aes(x = intSCT2_5, y = pred, group = ATG, col = ATG)) +
  geom_ribbon(
    aes(ymin = low, ymax = upp),
    fill = "gray", 
    alpha = 0.5, 
    col = NA
  ) +
  geom_line(size = 1.5, alpha = 0.75) +
  geom_label_repel(
    aes(label = label), 
    fontface = "bold",
    na.rm = TRUE, 
    hjust = 0,
    direction = "y",
    nudge_x = 0.5
  ) + 
  facet_wrap(facets = ~ VCMVPAT_pre, labeller = labeller(VCMVPAT_pre = cmv_labs)) +
  labs(x = "Time since alloHCT (months)", y = "CD4 cell counts") +
  scale_y_continuous(
    breaks = log(c(5, 25, 100, 500, 1500)),
    labels = c(5, 25, 100, 500, 1500)
  ) +
  theme_minimal() +
  theme(legend.position = "none") +
  coord_cartesian(xlim = c(0, 7.5)) +
  scale_color_brewer(palette = "Paired")
```

## NK joint model {.tabset}

### Specification

**Longitudinal sub-model:**

\begin{align*}
  \text{log(NK)} \sim \text{ATG} \times \text{Spline(Time)} + \text{CMV (Patient)},
\end{align*}

with $\text{Spline(Time)}$ being a flexible, natural *cubic* spline. Patient-specific spline parameters (three random effects)

**Competing risks sub-model:**

\begin{align*}
  \text{(1) DLI} &\sim \text{SCT (pre-2010)} + \text{Disease Risk}, \\
  \text{(2) Relapse} &\sim \text{ATG} + \text{Disease Risk} + \text{log(NK)}^*, \\
  \text{(3) Non-Relapse Failure: other} &\sim \text{ATG} + \text{log(NK)}^*, \\
  \text{(4) Non-Relapse Failure: GVHD} &\sim \text{log(NK)}^*, 
\end{align*}

with $\text{log(NK)}^*$ = "true" *current value* of log(NK) cell counts.

### Event submodel 

```{r nk_nointer_table}
summarise_event_submodel(NK_model_noInter) %>% 
  kableExtra::kbl(format = "html") %>%
  kable_paper("striped", full_width = F) %>%
  pack_rows("Cell. intervention", 1, 2) %>%
  pack_rows("Relapse", 3, 5) %>%
  pack_rows("NRF: Other", 6, 7) %>%
  pack_rows("NRF: GVHD", 8, 8)
```

### Predicted trajectories

```{r nk_traject}
preds_df_noInter[cell_type == "NK"] %>%
  ggplot(aes(x = intSCT2_5, y = pred, group = ATG, col = ATG)) +
  geom_ribbon(
    aes(ymin = low, ymax = upp),
    fill = "gray", 
    alpha = 0.5, 
    col = NA
  ) +
  geom_line(size = 1.5, alpha = 0.75) +
  geom_label_repel(
    aes(label = label), 
    fontface = "bold",
    na.rm = TRUE, 
    hjust = 0,
    direction = "y",
    nudge_x = 0.5
  ) + 
  facet_wrap(facets = ~ VCMVPAT_pre, labeller = labeller(VCMVPAT_pre = cmv_labs)) +
  labs(x = "Time since alloHCT (months)", y = "NK cell counts") +
  scale_y_continuous(
    breaks = log(c(5, 25, 100, 500, 1500)),
    labels = c(5, 25, 100, 500, 1500)
  ) +
  theme_minimal() +
  theme(legend.position = "none") +
  coord_cartesian(xlim = c(0, 7.5)) +
  scale_color_brewer(palette = "Paired")
```


## All cell trajectories

(Using models without interactions)

```{r all_traject}
preds_df_noInter %>%
  ggplot(
    aes(
      x = intSCT2_5,
      y = pred,
      ymin = low,
      ymax = upp,
      group = cell_type,
      col = cell_type
    )
  ) +
  geom_ribbon(fill = "gray", alpha = 0.5, col = NA) +
  geom_line(size = 1.5, alpha = 0.75) +
  facet_grid(
    facets = ATG ~ VCMVPAT_pre, 
    labeller = labeller(VCMVPAT_pre = cmv_labs, ATG = atg_labs)
  ) +
  labs(x = "Time since alloHCT (months)", y = "Cell counts") +
  scale_y_continuous(
    breaks = log(c(5, 25, 100, 500, 1500)),
    labels = c(5, 25, 100, 500, 1500)
  ) +
  theme_bw() +
  coord_cartesian(xlim = c(0, 6)) +
  scale_color_brewer("Cell type", palette = "Paired") +
  guides(colour = guide_legend(reverse = TRUE))
```

